---
title: "Machine Learning for Official Statistics and SDG Indicators "
author: "Your Name (Your Institution)"
subtitle: '*A simple report*'
date: "`r Sys.Date()`"
output:
  html_document:
    toc: yes
    toc_float: true
    code_folding: hide
  pdf_document:
    toc: yes
  word_document:
    toc: no
---



```{r libraries, include = FALSE}
# Load the required libraries

# For manipulating data
library(tidyverse)

# For tables display
library(flextable)

```


```{r setup, include=FALSE}
# Set echo=FALSE as default for code cells
# The code folding for HTLM output is thus deactivated
knitr::opts_chunk$set(echo = FALSE)

```


# Data analysis
We provide here some basic elements about the data set and descriptive statistics on some variables of interest.  

* First we load the data
* second we print a summary

```{r echo =FALSE}
TradeData <- read.csv("https://www.unsiap.or.jp/on_line/2024/RAP/Data/TradeData.csv")
```

We loaded the data set and compile the  report on `r Sys.Date()`

## Summary statistics
Here are some summary statistics of the export value (*Stat_Value* variable), all products combined:

```{r }
summary(TradeData$Stat_Value)
```

The data set contains   **`r  nrow(TradeData)`  observations**, and  **`r  ncol(TradeData)`** variables.

Here is a list of all variables available in the data set: 

```{r}
names(TradeData)
```

## Type of exports

The variable *Principle_Export* list all the types of goods exported from Vanuatu. There are **`r length(unique(TradeData$Principle_Exports))`** different goods exported. Let's have the list of all these goods. 

```{r}
unique(TradeData$Principle_Exports)

```

# Main Exports by type of goods

## As a table
It could be nicer to have this as a table:

```{r  }
# Calculate total value by principle export type
ExportsByType <- TradeData %>%
  group_by(Principle_Exports) %>%
  summarise(total = sum(Stat_Value)) %>%
  arrange(desc(total))

```



```{r}
# Table display
# Needs package "flextable"

# Printing the table
flextable(ExportsByType)
```

# As a graphic: 

Since we have a limited number of goods, a bar chart would probably describe the data in a very good way. We follow here some good practices in terms of data visualization: 

- The bar chart is **horizontal** to facilitate the reading of the labels for each good
- The bar chart is **ordered** to follow the data (largest on top)

```{r }
# Plot the exports
ExportsByType  %>%
  ggplot()+
  aes(x = reorder(Principle_Exports, total), y = (total)) +
  geom_bar(stat = "identity", 
             fill= "lightsteelblue1") +
  labs(title ="Exports from Vanuatu ", 
       subtitle = "Years [ 2000-2010]", 
       caption  = paste("Source: NSO, (file date: ",Sys.Date(), ")"), 
       x= "Exports categories",
       y = "Value (Vatu)") +
  coord_flip() +
  theme_minimal()

```

# Some extensions on this dynamic report

##  Selection of the Year

The previous analysis was done for all years. Imagine now the same report but for any year in [2000, 2010]. let's pick a year in the code chunck below! 

```{r}
MyYear <- 2003
```

> we have now selected the year **`r MyYear`**, let's see the ***new*** report!

### As a table

We use the same table as before. Maybe that could be a function later ;-).

```{r  }
# Now only for a specific year
ExportsByType_Year <- TradeData %>%
  filter(Year == MyYear) %>%
  group_by(Principle_Exports) %>%
  summarise(total = sum(Stat_Value)) %>%
  arrange(desc(total))

```


```{r}
flextable(ExportsByType_Year)
```


As expected, the graphic shows the difference with the 

```{r }
# Plot the exports for the specific year
ExportsByType_Year  %>%
  ggplot(aes(x = reorder(Principle_Exports, total), y = (total))) +  
    geom_bar(stat = "identity", color="white", fill= "lightsteelblue1") +
      labs( x= "Exports categories", 
            y = "Value (Vatu)", 
            title = paste("Exports from Vanuatu"),
            subtitle = paste(" Selected year:", MyYear), 
            caption  = paste("Source: NSO, (file date: ",Sys.Date(), ")"), ) +
  coord_flip() +
  theme_minimal()

```





